Renewal and stability in populations structured by remaining years of life

### 2015-11-06 ThanoRepro/ThanoRepro.tex
minor edits to full manuscript. Removed empirical section.

\begin{abstract}
A unisex model of population renewal for populations structured by
thanatological age is presented in both continuous and discrete form. Various
stable and transient properties populations are compared and related when viewed
by chronological versus thanatological age. Results from the two age perspectives are conformable if the starting
population is stable, but are otherwise divergent.\ackn{Research reported in this manuscript
was supported by the U.S.
National Institute On Aging of the National Institutes of Health under award numbers R01-AG011552 and R01-AG040245 and by the Spanish Ministry of Science and Innovation under award
number ECO2013-48326-C2-1-P. The content is solely the responsibility of the
author and does not necessarily represent the official views of the funding
agencies.}
\end{abstract}

@@

\onehalfspacing
\section*{Introduction}
In this paper, I explore some formal demographic consequences of a particular redefinition of age. Instead of counting age as the time passed since birth, consider age as the amount of time left until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure, or just time-to-death (TTD) elsewhere.}. The first
kind of age is called chronological age, and the second kind of age is called
thanatological age. Individuals in this case move in the same direction along
imaginary life lines, but the reference point for thanatological age is at
the end of the life line instead of at the beginning. For individuals, the timing of events in life is exactly identified using either of the two definitions of age, but
aggregate patterns differ. Demography is built upon a small set of empirical
regularities with respect to chronological age patterns. The models and methods
of demography are in varying degrees prefaced on such regularity. There is evidence that many demographic phenomena are best described in the
aggregatate as a function of time since birth, and others of time until death,
while others can be a function of both to some degree
\citep{riffe2015ttd}. This observation ought to motivate the formal demographic
base to incorporate this aspect of time in into models. 
%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new insights, questions, and conundrums.
% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

I describe the Lotka-Euler-Leslie renewal model when structured by
thanatological age. The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines of contemporary demographic research are relatable to the renewal model, and by extention to stable population theory.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here provide a broader treatment allowing for
growth in the population.
The chronological-age classified renewal model and the thanatological renewal model both describe the movement of population stocks over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox available to approach complex
temporal population structure. The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph.

\section*{A lifetable approximation of thanatological age}
%Direct information on the timing of demographic transitions with respect to
%death is available retrospectively from linked population registers or
%panel surveys with mortality follow-ups.
%Some surveys also ask respondents in various ways about their perceived future
% lifetime, which may underly those demographic transitions governed to some degree by volition. 
In general, to make a lifetable-based projective statement about the remaining
time distribution of a current stock of population, it is best to to account for population subgroups
with different mortality and to account for uncertainty about future changes in
mortality.
The present inquiry approximates thanatological age using lifetable-projected remaining lifetime,
which is intended to approximate remaining lifetime under the strong
assumptions of homogenous populations und fixed vital rates. 
%Chronological age-classified counts are redistributed over future death times
% according to a fixed mortality schedule.
%In general, a lifetable approach to transforming chronologically classified
%data into thanatologically classified data is sound if the given lifetable
%accurately describes the future mortality of the age-classified population, and
%if the chronological and thantological age patterns are independent.
In this case, each birth cohort undergoes the same
pattern of attrition. In a period setting, this means that we can distribute the
remaining lifespans of the population in each age class using the same
lifetable.
%This is the case we explore, and so the following formulas assume a single lifetable pattern.

The basic transformation to thanatological age proceeds by defining a
conditional deaths density function, $f(y|a)$. For chronological
age zero, $f(y|0)$ is identical to $d(a)$. For ages higher than 0, simply
condition on survival:\footnote{In the reliability literature, $f(y|a)$ is
denoted by $ X - t | X > t $, where $X$ is the total lifespan and $t$ is the
current chronological age attained.}
\begin{equation}
\label{eq:vaupel1}
f(y | a) = \mu(a+y)\frac{\ell(a+y)}{\ell(a)} \ep
\end{equation}

@@

Since $1 = \int_0^\infty f(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age,
such as the population pyramid, into thanatological age classes.
Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data \citep{HMD},
assuming the 2010 period mortality schedule remains constant. When looking from the chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this recomposition for the same 2010
US data.

@@

\citet{brouard1986structure} applied this method to highlight regional
differences in French age structure, at times combining this projective method
with historical stocks. \citet{brouard1989mouvements}, and later
\citet{vaupel2009life}, showed that in stationary populations the
marginal chronological and thanatological age structures are identical, and this
provides a limited basis for comparison between the two perspectives. 
%\citet{wachter2014essential} hints at, and
%\citet{vaupel2014stable} elaborates on how to make chronological and
%thanatological age structure commensurate under a constant growth rate, $r$.
These results will also follow from the following specification of a
thanatological renewal model.

It bears repeating that lifetable-derived thanatological age structure is
probabilistic until after death, since it refers to the future according to an
assumed mortality pattern; in this case a period mortality pattern is used.
As such, Figure~\ref{fig:USrecomp} is \textit{projective} in nature, a
forward-looking glance at population attrition given the population stock and lifetable of a
particular moment, whereas Figure~\ref{fig:USdecomp} is \textit{reflective} in
nature, since a population's chronological age-structure is mostly the fruit of past
fertility, but also migration, and to a lesser degree attrition. Any count
classified by chronological age can be reclassified in this way given a
corresponding lifetable. To derive thanatological fertility
rates under these assumptions, $\gamma(y)$, apply \eqref{eq:transform} to
chronological age-classified birth counts, $B(a)$, to get $B(y)$ and again to
exposure-to-risk, $E(a)$, to calcualte $E(y)$ and then divide in the usual way:

@@

Thanatological fertility rates, $\gamma(y)$, are briefly shown because they
are used in the renewal model here studied. Figure~\ref{fig:FySpaghetti}
represents the full variety in female thanatological period fertility rates that can be found for all years
of data that overlap in the HFD and HMD.\footnote{There are as of this writing
1933 population-years of overlap between the HMD and HFD, including a wide
variety of fertility and mortality combinations.} Figure~\ref{fig:FxSpaghetti} gives ASFR for the same
populations and years as a more familiar reference, but note that both scales
are different. The shape of thanatological fertility curves is speculative and
due to the above-mentioned strong assumptions. Nonetheless it is best not to
include such rates into a model blindly, so we offer this schematic overview of
some of their characteristics.

Period thanatological fertility rates estimated in this way have a
characteristic shape. The distribution is wider than that of chronological
fertility rates. The range of post-childbearing lifespans stretches from 
zero, due e.g., to maternal mortality, to at least 100.\footnote{For instance,
Jeanne Calment lived for another 100 years after giving birth to her daughter}
This spaghetti plot includes both fertility booms and busts, as well as some
mortality crises (1918, WWII). Some patterns in the data (not
shown) include that the left tail (a gauge of how orphan-prone a population is)
has tended to fall over time. All populations have shown an increase in both the
mean and mode of thanatological age at childbearing over time. Several
contemporary populations now show modes of over 60 years. Period thanatological
total fertility rates track standard period total fertility rates closely, but tend to be somewhat higher (not shown).
% both figures created in R/HFDspaghetti.R
\begin{figure}[h!]
	\caption{Chronological and thantological fertility rates, all 1933
	country-year combinations present in both the HFD and HMD.* Note different x
	and y scales.}
	\label{fig:Fxcompare}
	\begin{center}
	\makebox[\textwidth]{
	\begin{subfigure}{.45\textwidth}
		\caption{Chronological period fertility rates.}
		\label{fig:FxSpaghetti}
		\includegraphics[scale=.45]{Figures/FxSpaghettiDraft.png}
	\end{subfigure}
	~
	\begin{subfigure}{.45\textwidth}
		\caption{Thanatological period fertility rates.}
		\label{fig:FySpaghetti}
		\includegraphics[scale=.45]{Figures/FySpaghettiDraft.png}
	\end{subfigure}
	}
	\\
	\end{center}
	\begin{tiny}
	*AUT, $1951-2010$; BGR, $1947-2009$; BLR, $1964-2012$; CAN, $1921-2009$; 
	CHE, $1932-2011$; CHL, $1992-2005$; CZE, $1950-2011$; DEUTE, $1956-2011$; 
	DEUTNP, $1990-2011$; DEUTW, $1956-2011$; ESP, $1922-2012$; EST, $1959-2011$; 
	FIN, $1939-2009$; FRATNP, $1946-2013$; GBR\_NIR, $1974-2011$; GBR\_NP,
	$1974-2011$; GBR\_SCO, $1945-2011$; GBRTENW, $1938-2011$; HUN, $1950-2009$;
	IRL, $1955-2009$; ISL, $1963-2010$; JPN, $1947-2012$; LTU, $1959-2011$; NLD, $1950-2009$; 
	NOR, $1967-2009$; PRT, $1940-2012$; RUS, $1959-2010$; SVK, $1950-2009$; 
	SVN, $1983-2009$; SWE, $1891-2011$; TWN, $1976-2010$; UKR, $1959-2009$; 
	USA, $1933-2010$
	\end{tiny}
\end{figure}

@@

\FloatBarrier
\section*{The thanatological renewal model}
The renewal model that follows uses thanatological fertility rates, $\gamma(y)$,
the form of which is suggested in Figure~\ref{fig:FySpaghetti}, and we refer to
a homogenous, closed, unisex population based on female vital rates. Assuming constant vital rates, the births for the present year are given by 
\begin{align}
B(t) =& \int _0^\infty P(y,t)\gamma(y) \dd y = \int _0^\infty
P(a,t)m(a) \dd a \ec \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $\gamma(y), m(a)$ are
exact fertility probabilities (rates), best imagined
as the fertility rates of females born to females. The thanatological
(first) integral can be broken down back in terms of chronological age:} =&
\int_{y=0}^\infty \int_{a=0}^\infty \gamma(y)P(a,t)f(y|a) \dd a \dd y \ec
\intertext{where $f(y|a)$ is defined per equation~\eqref{eq:vaupel1}, so
$d(a)/\ell(0)$.
We can relate the present population to past births with $P(a,t) = B(t-a)\ell(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty \gamma(y) B(t-a)d(a+y)\dd a \dd y \ep 
\intertext{Eventually strong ergodicity will assert itself, and $B(t)$ will
be related to $B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty \gamma(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\ep
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty \gamma(y) d(a+y)e^{-ra}\dd a \dd y
\ep
\end{align}
Now compare this to Lotka's chronological formulation
\begin{align}
\label{eq:lotka}
1 =& \int_{a=0}^\infty m(a)l(a)e^{-ra}\dd a \ec
\intertext{which can be expressed as well as}
  =& \int_{a=0}^\infty \int_{y=0}^\infty m(a)d(a+y)e^{-ra}\dd y \dd a \ec
  \label{eq:lotka2}
\end{align}
and note that \eqref{eq:lotka2} and \eqref{eq:thanoren} are really quite
similar, since $\int _{y=0}^\infty d(a+y)\dd a = l(a)$. For intuition, notice
that $l(a)$ is here decomposed into $d(a)$, and imagine a matrix of these,
where one margin is chronological age and the other margin is thanatological age.
For the chronological case \eqref{eq:lotka}, we multiply chronological age-specific fertility rates, $f(a)$, over the chronological age margin, and for the thanatological case we multiply
$\gamma(y)$ over the thanatological age margin.

There are gaps in the above line of development, since the jump from
\eqref{eq:ergostep1} to \eqref{eq:ergostep2} (strong ergodicity) is assumed for
the moment. The step has some intuition, due to
the much greater density of connections within the model; persons from nearly any
thanatologcal age can produce offspring that can have any other
thanatological age.\footnote{I say \textit{nearly} because the very highest thanatological ages tend to be composed of pre-menarchical females, so these can be thought of as structural
zeros.} In
this sense, the smoothing mechanism at play as the population passes through
time must be much stronger than that for chronological age, at least in most
cases.  The mechanisms at play unfold in the same way as those so intuitively
described by \citet{arthur1982ergodic}, and said proof may apply here without
further modification. Later in this paper I note how strong ergodicity is
guaranteed given the properties of the thanatolgical projection matrix. A proof
of the uniqueness of the solution to \eqref{eq:thanoren} is given in
Appendix~\ref{app:A}. A proof that the chronological and thanatological renewal
models must yield the same $r$ if the starting population is already stable is
given in Appendix~\ref{app:B}. A fast-converging iterative method for finding
$r$ from equation~\eqref{eq:thanoren} is given in Appendix~\ref{app:C}

@@

Once one finds $r$ from \eqref{eq:thanoren}, other familiar stable population
parameters can be calculated. For instance, we may calculate the mean
thanatological generation time, $T^\star$:
\begin{equation}
\label{eq:Ty}
 T^\star =  \frac{\int _{y=0}^\infty \int _{a=0}^\infty y e^{-ra} d(a+y)
 \gamma(y) \dd a \dd y}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y)
 \gamma(y) \dd a \dd y} \ep
\end{equation}
Literally, this is the mean of the remaining lifespans of new mothers in the
stable population. The net reproduction rate, $R_0$ is related by, e.g.,
\begin{equation}
\label{eq:R0fromTy}
R_0 = e^{r T^\star}
\end{equation}
The birth rate, $b$, is given by
\begin{align}
\label{eq:eybrate}
b =& \frac{1}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y) \dd a \dd
y}\ec
\intertext{the denominator of which is the inside of
equation~\eqref{eq:thanoren} less $\gamma(y)$. This simplifies to} =&
\frac{1}{ \int _{a=0}^\infty e^{-ra} l(a) \dd a } \ec
\intertext{which is exactly the same as that valid for
chronological age in the Lotka-Euler setup, except that $r$ here comes from the
thanatological renewal model. If the stable age structure, $c^\star(y)$, is
known, then we may take the sum of hypothetical births over thanatological age:}
=& \int _{y=0}^\infty \gamma(y) c^\star(y) \dd y \ep
\end{align}

The stable thanatological age structure, $c^\star(y)$, is the proportion of the stable population with remaining years to live $y$
\begin{align}
\label{eq:cy}
c^\star(y) =& b \int _{a=0}^\infty e^{-ra} d(a+y) \dd a \ec
\intertext{or, given the stable chronological age structure, $c(a)$, and a
lifetable,} =& \int _{a=0}^\infty c(a)d(y|a) \dd a \label{eq:cyfromca}\\ \notag
\\
           =& \frac{\int _{a=0}^\infty d(a+y)e^{-ra}\dd a}{\int
           _{a=0}^\infty l(a)e^{-ra}\dd a}  \ep
\end{align}

@@

\section*{The thanatological projection matrix}
These descriptions can be made more explicit, and in some ways more tractable,
by hashing out the projection matrix that corresponds to the thanatological renewal model. As with the
chronological age-structured Leslie matrix, the thanatological projection
matrix, $\textbf{Y}$, is square and of dimension $y \times y$, where $y$ is the number
of thanatological age classifications on which the population is structured. 
The matrix contains elements for survival and elements for fertility. Unlike
Leslie matrices, $\textbf{Y}$ is not sparse, but is instead populated primarily
with positive entries greater than zero. In the
following, I illustrate using a 6$\times$6 matrix.

>>>>

\begin{abstract}
A unisex model of population renewal for populations structured by
remaining years of life is presented in both continuous and discrete form.
Various stable and transient properties populations are compared and related when viewed
by chronological versus thanatological age. Results from the two age perspectives are conformable if the starting
population is stable, but are otherwise divergent.\ackn{Research reported in this manuscript
was supported by the U.S.
National Institute On Aging of the National Institutes of Health under award numbers R01-AG011552 and R01-AG040245 and by the Spanish Ministry of Science and Innovation under award
number ECO2013-48326-C2-1-P. The content is solely the responsibility of the
author and does not necessarily represent the official views of the funding
agencies.}
\end{abstract}

@@

\onehalfspacing
\section*{Introduction}
In this paper, I explore some formal demographic consequences of a particular redefinition of age. Instead of counting age as the time passed since birth, consider age as the amount of time left until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure, or just time-to-death (TTD) elsewhere.}. The first
kind of age is called chronological age, and the second kind of age is called
thanatological age. Individuals in this case move in the same direction along
imaginary life lines, but the reference point for thanatological age is at
the end of the life line instead of at the beginning. For individuals, the timing of events in life is exactly identified using either of the two definitions of age, but
aggregate patterns differ. Demography is built upon a small set of empirical
regularities with respect to chronological age patterns. The models and methods
of demography are in varying degrees prefaced on such regularity. There is evidence that many demographic phenomena are best described in the
aggregatate as a function of time since birth, and others of time until death,
while others can be a function of both to some degree
\citep{riffe2015ttd,wolf2015disability}. This observation ought to motivate the
formal demographic base to incorporate this aspect of time in into models. 
%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new
% insights, questions, and conundrums.
% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

I describe the Lotka-Euler-Leslie renewal model when structured by
thanatological age. The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines of contemporary demographic research are relatable to the renewal model, and by extention to stable population theory.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here provide a broader treatment allowing for
growth in the population.
The chronological-age classified renewal model and the thanatological renewal model both describe the movement of population stocks over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox available to approach complex
temporal population structure.

The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph.

\section*{A lifetable approximation of thanatological age}
%Direct information on the timing of demographic transitions with respect to
%death is available retrospectively from linked population registers or
%panel surveys with mortality follow-ups.
%Some surveys also ask respondents in various ways about their perceived future
% lifetime, which may underly those demographic transitions governed to some degree by volition. 
In general, to make a lifetable-based projective statement about the remaining
time distribution of a current stock of population, it is best to to account for population subgroups
with different mortality and to account for uncertainty about future changes in
mortality.
The present inquiry approximates thanatological age using lifetable-projected remaining lifetime,
which is intended to approximate remaining lifetime under the strong
assumptions of homogenous populations und fixed vital rates. 
%Chronological age-classified counts are redistributed over future death times
% according to a fixed mortality schedule.
%In general, a lifetable approach to transforming chronologically classified
%data into thanatologically classified data is sound if the given lifetable
%accurately describes the future mortality of the age-classified population, and
%if the chronological and thantological age patterns are independent.
Under these assumptions, each birth cohort undergoes the same
pattern of attrition. In a period setting, this means that we can distribute the
remaining lifespans of the population in each age class using the same
lifetable.
%This is the case we explore, and so the following formulas assume a single lifetable pattern.

The basic transformation to thanatological age proceeds by defining a
conditional deaths density function, $f(y|a)$. For chronological
age zero, $f(y|0)$ is identical to $d(a)$, the lifetable deaths distribution
(based on a unity radix).
For ages higher than 0, simply condition on survival:\footnote{In the reliability literature, $f(y|a)$ is
denoted by $ X - t | X > t $, where $X$ is the total lifespan and $t$ is the
current chronological age attained.}
\begin{equation}
\label{eq:vaupel1}
f(y | a) = \mu(a+y)\frac{\ell(a+y)}{\ell(a)} \ep
\end{equation}

@@

Since $1 = \int_0^\infty f(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age,
such as the population pyramid, into thanatological age classes.
Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data \citep{HMD},
assuming the 2010 period mortality schedule remains constant. When looking from the chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this recomposition for the same 2010
US data. The outer profile of Figure~\ref{fig:USrecomp} is $P(y)$, under our assumptions.

@@

\citet{brouard1986structure} applied this method to highlight regional
differences in French age structure, at times combining this projective method
with historical stocks. \citet{brouard1989mouvements}, and \citet{vaupel2009life}, showed that in stationary populations the
marginal chronological and thanatological age structures are identical, and this
provides a limited basis for comparison between the two perspectives. 
%\citet{wachter2014essential} hints at, and
%\citet{vaupel2014stable} elaborates on how to make chronological and
%thanatological age structure commensurate under a constant growth rate, $r$.
These results also follow from the following specification of a
thanatological renewal model.

Lifetable-derived thanatological age structure is
probabilistic until after death, since it refers to the future according to an
assumed mortality pattern; in this case a period mortality pattern is used.
As such, Figure~\ref{fig:USrecomp} is \textit{projective} in nature, a
forward-looking glance at population attrition given the population stock and lifetable of a
particular moment, whereas Figure~\ref{fig:USdecomp} is \textit{reflective} in
nature, since a population's chronological age-structure is mostly the fruit of past
fertility, but also to a lesser degree migration and attrition. Any count
classified by chronological age can be reclassified in this way given a
pertinent lifetable. To
derive thanatological fertility rates under these assumptions, $\gamma(y)$, apply \eqref{eq:transform} to
chronological age-classified birth counts, $B(a)$, to get $B(y)$ and again to
exposure-to-risk, $E(a)$, to calcualte $E(y)$ and then divide in the usual way:

@@

Figure~\ref{fig:FySpaghetti}
represents the full variety in female thanatological period fertility rates that can be found for all years
of data that overlap in the HFD and HMD.\footnote{There are as of this writing
1933 population-years of overlap between the HMD and HFD, including a wide
variety of fertility and mortality combinations.} I show $\gamma(y)$ calculated in this way because
it enters into the renewal model here studied. Figure~\ref{fig:FxSpaghetti}
gives ASFR for the same populations and years as a more familiar reference, but
note that both the x and y axes are different. The shape of thanatological
fertility curves is speculative and due to the above-mentioned strong assumptions. Nonetheless it is best not to
include such rates into a model blindly, so we offer this schematic overview of
some of their characteristics.

Period thanatological fertility rates estimated in this way have a
characteristic shape. The distribution is wider than that of chronological
fertility rates. The range of post-childbearing lifespans stretches from 
zero, due e.g., to maternal mortality, to at least 100.\footnote{For instance,
Jeanne Calment lived for another 100 years after giving birth to her daughter.}
This spaghetti plot includes both fertility booms and busts, as well as some
mortality crises (1918, WWII). Some patterns in the data (not
highlighted in the figure) include that the left tail (a gauge of how
orphan-prone a population is) has tended to fall over time. All populations have shown an increase in both the
mean and mode of thanatological age at childbearing over time. Several
contemporary populations now show modes of over 60 years. Period thanatological total
fertility rates track standard period total fertility rates closely, but tend to be somewhat higher (not shown).
% both figures created in R/HFDspaghetti.R
\begin{figure}[h!]
	\caption{Chronological and thantological fertility rates, all 1933
	country-year combinations present in both the HFD and HMD.* Note different x
	and y scales.}
	\label{fig:Fxcompare}
	\begin{center}
	\makebox[\textwidth]{
	\begin{subfigure}{.45\textwidth}
		\caption{Chronological period fertility rates.}
		\label{fig:FxSpaghetti}
		\includegraphics[scale=.45]{Figures/FxSpaghettiDraft.png}
	\end{subfigure}
	~
	\begin{subfigure}{.45\textwidth}
		\caption{Thanatological period fertility rates.}
		\label{fig:FySpaghetti}
		\includegraphics[scale=.45]{Figures/FySpaghettiDraft.png}
	\end{subfigure}
	}
	\\
	\end{center}
	\begin{tiny}
	*AUT, $1951-2010$; BGR, $1947-2009$; BLR, $1964-2012$; CAN, $1921-2009$; 
	CHE, $1932-2011$; CHL, $1992-2005$; CZE, $1950-2011$; DEUTE, $1956-2011$; 
	DEUTNP, $1990-2011$; DEUTW, $1956-2011$; ESP, $1922-2012$; EST, $1959-2011$; 
	FIN, $1939-2009$; FRATNP, $1946-2013$; GBR\_NIR, $1974-2011$; GBR\_NP,
	$1974-2011$; GBR\_SCO, $1945-2011$; GBRTENW, $1938-2011$; HUN, $1950-2009$;
	IRL, $1955-2009$; ISL, $1963-2010$; JPN, $1947-2012$; LTU, $1959-2011$; NLD, $1950-2009$; 
	NOR, $1967-2009$; PRT, $1940-2012$; RUS, $1959-2010$; SVK, $1950-2009$; 
	SVN, $1983-2009$; SWE, $1891-2011$; TWN, $1976-2010$; UKR, $1959-2009$; 
	USA, $1933-2010$
	\end{tiny}
\end{figure}

@@

\FloatBarrier
\section*{The thanatological renewal model}
The renewal model that follows uses thanatological fertility rates, $\gamma(y)$,
the form of which is suggested in Figure~\ref{fig:FySpaghetti}, and we refer to
a homogenous, closed, unisex population based on female vital rates. Assuming constant vital rates, the births for the present year are given by 
\begin{align}
B(t) =& \int _0^\infty P(y,t)\gamma(y) \dd y = \int _0^\infty
P(a,t)m(a) \dd a \ec \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $\gamma(y), m(a)$ are
exact fertility probabilities (rates), best imagined
as the fertility rates of females born to females. The thanatological
(first) integral can be broken down back in terms of chronological age:} =&
\int_{y=0}^\infty \int_{a=0}^\infty \gamma(y)P(a,t)f(y|a) \dd a \dd y \ec
\intertext{where $f(y|a)$ is defined per equation~\eqref{eq:vaupel1}, so
$d(a)/\ell(0)$.
We can relate the present population to past births with $P(a,t) = B(t-a)\ell(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty \gamma(y) B(t-a)d(a+y)\dd a \dd y \ep 
\intertext{Eventually strong ergodicity will assert itself, and $B(t)$ will
be related to $B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty \gamma(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\ep
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty \gamma(y) d(a+y)e^{-ra}\dd a \dd y
\ep
\end{align}
Now compare this to Lotka's chronological formulation:
\begin{align}
\label{eq:lotka}
1 =& \int_{a=0}^\infty m(a)l(a)e^{-ra}\dd a \ec
\intertext{which can be expressed as well as}
  =& \int_{a=0}^\infty \int_{y=0}^\infty m(a)d(a+y)e^{-ra}\dd y \dd a \ec
  \label{eq:lotka2}
\end{align}
and note that \eqref{eq:lotka2} and \eqref{eq:thanoren} are really quite
similar, since $\int _{y=0}^\infty d(a+y)\dd a = l(a)$. For intuition, notice
that $l(a)$ is here decomposed into $d(a)$. Imagine a matrix of this
decomposition, where one margin is chronological age and the other margin is
thanatological age.
For the chronological case \eqref{eq:lotka}, we multiply chronological
age-specific fertility rates, $m(a)$, over the chronological age margin, and for the thanatological case we multiply $\gamma(y)$ over the thanatological age margin.

There are gaps in the above line of development, since the jump from
\eqref{eq:ergostep1} to \eqref{eq:ergostep2} (strong ergodicity) is assumed for
the moment.\footnote{The step to exponential stability is proven
for the discrete version of the model.} The step has some intuition, due to
the much greater density of connections within the model; persons from nearly any
thanatologcal age can produce offspring that can have any other
thanatological age.\footnote{I say \textit{nearly} because the very highest thanatological ages tend to be composed of pre-menarchical females, so these can be thought of as structural
zeros.} In
this sense, the smoothing mechanism at play as the population passes through
time must be much stronger than that for chronological age, at least in most
cases. The mechanisms at play unfold in the same way
as those so intuitively described by \citet{arthur1982ergodic}, and said proof may apply here without
further modification. Later in this paper I note how strong ergodicity is
guaranteed given the properties of the thanatolgical projection matrix. A proof
of the uniqueness of the solution to \eqref{eq:thanoren} is given in
Appendix~\ref{app:A}. A proof that the chronological and thanatological renewal
models must yield the same $r$ if the starting population is already stable is
given in Appendix~\ref{app:B}. A fast-converging iterative method for finding
$r$ from equation~\eqref{eq:thanoren} is given in Appendix~\ref{app:C}

@@

Once one finds $r$ from \eqref{eq:thanoren}, other familiar stable population
parameters can be calculated. For instance, we may calculate the mean
thanatological generation time, $T^\star$:
\begin{equation}
\label{eq:Ty}
 T^\star =  \frac{\int _{y=0}^\infty \int _{a=0}^\infty y e^{-ra} d(a+y)
 \gamma(y) \dd a \dd y}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y)
 \gamma(y) \dd a \dd y} \ep
\end{equation}
Literally, this is the mean of the remaining lifespans of new mothers in the
stable population. The thanatological net reproduction rate, $R_0^\star$ is
related by, e.g.,
\begin{equation}
\label{eq:R0fromTy}
R_0^\star = e^{r T^\star}
\end{equation}
The birth rate, $b$, is given by
\begin{align}
\label{eq:eybrate}
b =& \frac{1}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y) \dd a \dd
y}\ec
\intertext{the denominator of which is the inside of
equation~\eqref{eq:thanoren} less $\gamma(y)$. This simplifies to} =&
\frac{1}{ \int _{a=0}^\infty e^{-ra} l(a) \dd a } \ec
\intertext{which is exactly the same as that valid for
chronological age in the Lotka-Euler setup, except that $r$ here comes from the
thanatological renewal model. If the stable age structure, $c^\star(y)$, is
known, then we may take the sum of hypothetical births over thanatological age:}
=& \int _{y=0}^\infty \gamma(y) c^\star(y) \dd y \ep
\end{align}

The stable thanatological age structure, $c^\star(y)$, is the proportion of the stable population with remaining years to live $y$
\begin{align}
\label{eq:cy}
c^\star(y) =& b \int _{a=0}^\infty e^{-ra} d(a+y) \dd a \ec
\intertext{or, given the stable chronological age structure, $c(a)$, and a
lifetable,} =& \int _{a=0}^\infty c(a)f(y|a) \dd a \label{eq:cyfromca}\\ \notag
\\
           =& \frac{\int _{a=0}^\infty d(a+y)e^{-ra}\dd a}{\int
           _{a=0}^\infty l(a)e^{-ra}\dd a}  \ep
\end{align}

@@

\section*{The thanatological projection matrix}
These descriptions can be made more explicit, and in some ways more tractable,
by hashing out the projection matrix that corresponds to the thanatological renewal model. As with the
chronological age-structured Leslie matrix, the thanatological projection
matrix, $\textbf{Y}$, is square and of dimension $y \times y$, where $y$ is the number
of equal-interval thanatological age classes into which the population is
structured.
The matrix contains elements for survival and elements for fertility. Unlike
Leslie matrices, $\textbf{Y}$ is not sparse, but is populated primarily
with positive entries greater than zero. In the
following, I illustrate using a 6$\times$6 matrix.
